Darwin's Dream: Unifying ecological and evolutionary change

### 2025-05-29 response_to_reviewers.md

What is "10.5" exactly? It has no meaning, as it is a unitless composition of both the number of individuals and the sum of the traits they possess. On Line 192, the authors suggest it might be £a quantity of ecosystem function", but I can’t imagine how that could be the case given that what it encompasses (traits and individuals) changes depending on the values of the summation, and different values of traits, birth, and death, would produce identical Omega values.

Author response to Reviewer 2 comment 2
=====================

As Reviewer 2 correctly states, the quantity Omega is abstract and lacks a concrete interpretation or predefined units in eqn 1. This isn't a problem and is actually necessary given what we are trying to achieve, which is to identify the equation from which fundamental equations of ecology and evolution can be logically derived. This requires a very high level of abstraction to encompass any possible eco-evolutionary change.

The variable Omega is abstract and unitless for the exact same reason that 'z' is abstract and unitless in the Price equation. The Price equation requires a very high level of abstraction to be able to describe the change in some characteristic z for any possible evolving system. Units of z could therefore reflect any evolving characteristic (e.g., body size volume, discrete behaviour, phenology, number of hairs on a wing, carbon capture -- any measurable characteristic can be assigned to 'z' in the Price equation). Units of Omega in our equation will be identical to units of z, as used in the Price equation. Omega is just the sum across all individuals.

We appreciate that a concrete example can be helpful, and we thank Reviewer 2 for the suggestion. Here is how we would meaningfully use it: Let this small population of N = 3 be three different plants, which Reviewer 2 has assigned individual one a fecundity of beta = 2 and individuals two and three a fecundity of beta = 1. This population might be interpreted to have non-overlapping generations because death occurs for all individuals (delta = 1). Assume that we are interested in the rate of carbon capture of these plants, so this is the characteristic 'z' that we are measuring (e.g., in kg of carbon dioxide absorbed over some period of time). The DeltaZ = 0.5 would be a bit unusual, but could be caused by a change in carbon capture from t to t + 1 due to changing environmental conditions affecting the plants.

We could then calculate Omega as the summed rate of carbon capture attributable to the population from t to t + 1.

Omega = (2 - 1 + 1)(1 + 0.5) + (1 - 1 + 1)(2 + 0.5) + (1 - 1 + 1)(3 + 0.5) = 9

Note that Omega = 9 in the example Reviewer 2 gave (the 10.5 was just a calculation error). That is, the total carbon capture summed across all individuals is 9 kg. There is no problem with units here; the correct unit is kg, not a combination of individuals and kg.

Counts are unitless, inherently, and 'individual' is a label rather than a unit in this case, regardless of whether it is in beta (births attributable to individual i), delta (death indicator of i), or 1 (the count of i itself).

From the Bureau International des Poids et Mesures (2019) The International System of Units (SI) (9th ed.) (https://www.bipm.org/en/publications/si-brochure/): "There are also some quantities that cannot be described in terms of the seven base quantities of the SI, but are quantities that are a number of entities. Examples are a number of cellular or biomolecular entities, or degeneracy in quantum mechanics. These quantities are also quantities with the unit one", then "The unit one is the neutral element of any system of units – necessary and present automatically. There is no requirement to introduce it formally by decision."

A major part of our work here is to set the fundamental equations of ecology and evolution on as solid of a foundation as possible. And while the point about units that Reviewer 2 raises is thought-provoking (we had to go back to the basics ourselves to work out whether or not this was an issue for our equation), the logic of our equation remains airtight.

Intuitively, note that if we were just interested in measuring total carbon capture in a population, then it wouldn't be a problem to, hypothetically, just sum up how much each individual captures carbon. Similarly, if we wanted to measure total biomass, or decomposition, or pollination in a system (anything, really), it would be entirely sensible (if it were possible in practice) to add up the contributions of each individual to the whole. And this is what we're after with Omega -- the summed value of traits across all individuals.

Now, in the special case in which z = 1 (which is, again, a unitless count), we recover the total number of individuals, as derived in our manuscript. In this case, Omega is interpreted as individual number (N), and,

Omega = (2 - 1 + 1)(1 + 0) + (1 - 1 + 1)(1 + 0) + (1 - 1 + 1)(1 + 0) = 4

Our population size has increased from 3 to 4 because two individuals produced one offspring each (beta = 1), and one produced two (beta = 2). All of those reproducing individuals died (delta = 1).

We thank Reviewer 2 for raising this query; it gave us pause to reflect and be even more deliberate about what our equation is describing. We now mention in our manuscript why this is not an issue by pointing out that individuals are unitless (LINE).

Reviewer 2 comment 3 
======================

Thus, if I'm understanding correctly, equation 1 only becomes meaningful once the steps are taken to reduce it to either the population growth rate equation or the Price Equation. This feels like deriving a fundamental equation in reverse – starting from two known fundamental equations, and then working backwards to piece them together. Which perhaps is a fine approach on its face, but if this equation does not produce evolutionarily or ecologically meaningful results on its own, it's unclear that it has any utility.

Author response to Reviewer 2 comment 3
======================

This is incorrect; eqn 1 is meaningful as written as a definition of eco-evolutionary change. It is, however, correct that the units associated with 'z' (and therefore 'Omega') are unresolved until 'z' is defined (see our response to Reviewer 2 comment 2 above). But this is no different from the Price equation itself. Our equation doesn't need to recover either the population growth rate equation or the Price Equation to be meaningful. On the contrary, the logic of our manuscript is to develop a unifying framework for eco-evolutionary change. Of course it is good and necessary to prove that we can recover the two most fundamental equations of ecology and evolution. But that is not the end of the story. If our equation only worked through specifying an ecological scenario or an evolutionary scenario, but not both, then it would not fulfil its intended purpose.

We can appreciate why it might feel like our equation 1 works backwards from the fundamental equations of ecology and evolution separately. But this isn't the case, logically or methodologically (see our response to Reviewer 2 comment 1). What we show is that the appropriate starting point for defining eco-evolutionary change is the interaction between births/deaths and individual characteristics. Our starting point isn't arbitrary or contingent; it is exact and universal, and no other equation would work as fundamental (anymore than the Price equation or general population change could be substituted by something else as fundamental equations).

We now emphasise more clearly, in words, the interpretation of equation 1 and how it is meaningful even prior to the derivation of population change or the Price equation (LINE XXX).

Reviewer 2 comment 4 
=========================

Lastly, there are a series of statements made in the Abstract that feel like they either are not followed through on or only weakly mentioned in the Discussion. For example, we are told that this unification “shows how ecological and evolutionary change are reflected in the first and second statistical moments of fitness”, but this is fleshed out in a single sentence in the Discussion (Line 236–238): “Interestingly, the rate of change in ecology and evolution are reflected in the first and second statistical moments of fitness, respectively. Population growth rate reflects mean fitness w ̄ , while the rate of evolutionary change reflects the variance in fitness Var(w).” There appears to be no other mention of this outside the Abstract. Importantly, as the authors note, each of these things are already known – both the equivalence of population growth rate with mean fitness and the rate of evolutionary change being equal to the variance in fitness are found in Fisher (1930).

Author response to Reviewer 2 comment 4 
=========================

We now remove this from the abstract. In the manuscript, we now clarify that our derivation recovers this relationship, which has been pointed out previously.

Reviewer 2 comment 5
=========================

Furthermore, the Abstract claims the unification links ecosystem function to population growth rates and evolutionary change, but this is only speculated on in the last paragraph of the Discussion. The authors cite others who have explicitly done this, suggest that their equation might do the same, but there’s no demonstration of it or suggestion of what it might be. Again, this could be because it’s not obvious to me the connection, so the authors need to be a little more explicit in showing these connections.

Author response to Reviewer 2 comment 5
=========================

We are grateful to Reviewer 2 for pointing this out, and we have revised to make our point about ecosystem function clearer with a more concise explanation of our point in our response to Reviewer 2 comment 2. We now elaborate on how Omega can be interpreted as a measure of ecosystem function for a population (LINE XXX).

Reviewer 2 minor comment 1
=========================

Title – I like the title, but Darwin is never mentioned in the text! I kept expecting to see something about how “Darwin dreamed to unite these fields” or something.

Author response to Reviewer 2 minor comment 1
=========================

We also liked the title, but it is clearly causing confusion. We have switched to 'Foundations of ecological and evolutionary change'.

Reviewer 2 minor comment 2
=========================

Lines 26-28 – I would also cite Betty Smocovitis’ excellent paper here on the unification of biology via evolution: Smocovitis, V. B. (1992). Unifying biology: the evolutionary synthesis and evolutionary biology. Journal of the History of Biology, 25(1), 1-65.

Author response to Reviewer 2 minor comment 2
=========================

This is a good suggestion; we have cited this paper now (Line XXX).

Reviewer 2 minor comment 3
=========================

Lines 45-54 – This historical presentation seems odd to me. My reading of history is that these two fields have always been pretty tightly intertwined; Darwin’s argument relied on Malthusian concepts of population growth, after all, and one of the first concepts of density dependent selection comes from Haldane (1956). But I could be biased. Perhaps a citation after the sentence: “Until recently, population ecology and evolution in particular have taken almost parallel paths for developing theory” to others who have argued this would be helpful.

Author response to Reviewer 2 minor comment 3 
=========================

We are not historians and we don’t claim to have the ultimate answer on this issue. Our reading of history is the following. The theory of evolution by natural selection developed by Darwin is inherently an ecological theory, and the last third of the 19th century saw the development of ecological work rooted on ivolution. In a work devoted to the historical relationship between ecology and evolution, James P. Collins (1986, p. 260) states: "In 1866, for example, Haeckel claimed that the theory of evolution explained and formed the groundwork of ecology. Stephen Forbes argued in 1895 that the whole Darwinian doctrine belonged to ecology. Just after 1900 H. C. Cowles wrote, "If ecology has a place at all in modern biology, certainly one of its great tasks is to unravel the mysteries of adaptation." At that same time, J. G. Needham described ecology as "the study of the phenomena of fitness."

However, that doesn’t imply that ecological theory/practice and evolutionary theory/practice have always been as intertwined as one would think. The progressive specialisation of both areas produced subdisciplines with no evolutionary input/perspective, or evolution (specially adaptation) was just taken for granted, orienting their research to mechanistic/proximate questions (in terms of Mayr and Tinbergen conceptualization of proximate vs. ultimate questions). Moreover, fields such as systems ecology (from Odum to Margalef) could be thought of as a renewed version of the Romantische Naturphilosophie, with little (if any) connection with evolution. Collins (1986, p. 258) cites works from the 1950s claiming for the necessity of an autonomous ecological theory, free of evolutionary thinking. All of this may explain remarks like the following during the 1960s: "The theory of evolution by natural selection is an ecological theory -founded on ecological observation by perhaps the greatest of all ecologists. It has been adopted by and brought up by the science of genetics, and ecologists, being modest people, are apt to forget their distinguished parenthood. Indeed, Darwinian plant ecology has been largely neglected" (Harper 1967 p. 267).

After 1960, evolutionary ecology emerged and many ecology textbooks (Pianka 1974, Roughgarden 1979, Begon et al. 1986/2006) included an evolutionary perspective and entire chapters devoted to evolution. However, we think that the pursuit of a true synthesis of evolutionary and ecological change (eco-evo dynamics) has only taken place (roughly) in the last two decades, with several works trying to develop a more unified framework (as far as we know, only one textbook (Hendry 2017) has summarized these eco-evo efforts). Part of the motivation for this synthesis has been the relatively recent appreciation of concurrent ecological and evolutionary dynamics. From Yamamichi, Ellner, and Hairston Jr. (2023), "Although ecology and evolutionary biology have been neighbouring research areas ever since the age of Darwin, the prevailing assumption had long been that ecological processes occur much faster than evolutionary processes".

One of our critiques of eco-evolutionary theory is that there is no formal integration of the foundational axioms shared by both ecology and evolution. Our take is that we should have an eco-evo formalism tout court, which takes a separate but complementary role to existing eco-evolutionary models, which require simplifying assumptions and typically start with ecology or with evolution and then try to include the other one in the model.

Finally, we also agree with Reviewer 2 that classic evolutionary works (like those developed by Haldane) had ecological ingredients. Actually, as we said in our manuscript, our claim is that, contrary to a spread view among evolutionary biologists (see page XXX of our manuscript for references), population genetics has ecology at its core, though not in an obvious way.

We now clarify this historical view (Line XXX)

Reviewer 2 minor comment 4
=========================

Line 106 – “Fischer” should be spelled “Fisher”.

Author response to Reviewer 2 minor comment 4
=========================

Fixed (Line XXX).

Reviewer 2 minor comment 5 
=========================

Lines 273-274 – If these derivations were included here, it would dramatically clarify my confusion about the usefulness and meaning of Omega in the general formula.

Author response to Reviewer 2 minor comment 5  
=========================

We are reluctant to include too much work still in preparation, or distract from what we believe is the central message of the current manuscript. But we have now included some additional derivations in supporting information.

Author references
==============================

Baravalle, L., Jonathan Roffé, A., Luque, V. J., & Ginnobili, S. (2024). The Value of Price. Biological Theory. https://doi.org/10.1007/s13752-024-00482-4

Barfield, M., Holt, R. D., & Gomulkiewicz, R. (2011). Evolution in stage-structured populations. American Naturalist, 177(4), 397–409. https://doi.org/10.1086/658903

Bureau international Des poids et mesures (2019) The International System of Units (SI).
https://www.bipm.org/documents/20126/41483022/SI-Brochure-9.pdf

Caswell, H. (1988). Theory and models in ecology: a different perspective. Ecological Modelling, 43, 33–44.

Coulson, T., & Tuljapurkar, S. (2008). The dynamics of a quantitative trait in an age-structured population living in a variable environment. American Naturalist, 172(5), 599–612. https://doi.org/10.1086/591693

Levins, R. (1966). The strategy of model building in population biology. In American Naturalist (Vol. 54, Issue 4, pp. 421–431). https://doi.org/10.2307/27836590

Frank, S. A. (1995). George Price’s contributions to evolutionary genetics. Journal of Theoretical Biology, 175, 373–388.

Frank, S. A. (1997). The Price equation, Fisher’s fundamental theorem, kin selection, and causal analysis. Evolution, 51(6), 1712–1729.

Frank, S. A. (2012). Natural selection. IV. The Price equation. Journal of Evolutionary Biology, 25, 1002–1019. https://doi.org/10.1111/j.1420-9101.2012.02498.x

Frank, S. A. (2015). D’Alembert’s direct and inertial forces acting on populations: The price equation and the fundamental theorem of natural selection. Entropy, 17(10), 7087–7100. https://doi.org/10.3390/e17107087

Frank, S. A. (2016). Common probability patterns arise from simple invariances. Entropy, 18(5), 1–22. https://doi.org/10.3390/e18050192

Price, G. R. (1970). Selection and covariance. In Nature (Vol. 227, Issue 5257, pp. 520–521). https://doi.org/10.1038/227520a0

Price, G. R. (1972). Extension of covariance selection mathematics. Annals of Human Genetics, 35(4), 485–490. https://doi.org/10.1111/j.1469-1809.1957.tb01874.x

Queller, D. C. (2017). Fundamental theorems of evolution. American Naturalist, 189(4), 000–000. https://doi.org/10.1086/690937

Lion, S. (2018). Theoretical approaches in evolutionary ecology: environmental feedback as a unifying perspective. American Naturalist, 191(1). https://doi.org/10.1086/694865

Lion, S., Sasaki, A., & Boots, M. (2023). Extending eco-evolutionary theory with oligomorphic dynamics. Ecology Letters, 26(S1), S22–S46. https://doi.org/10.1111/ele.14183

Patel, S., Cortez, M. H., & Schreiber, S. J. (2018). Partitioning the effects of eco-evolutionary feedbacks on community stability. American Naturalist, 191(3), 1–29. https://doi.org/10.1101/104505

Govaert, L., Fronhofer, E. A., Lion, S., Eizaguirre, C., Bonte, D., Egas, M., Hendry, A. P., de Brito Martins, A., Melián, C. J., Raeymaekers, J. A., Ratikainen, I. I., Saether, B. E., Schweitzer, J. A., & Matthews, B. (2019). Eco-evolutionary feedbacks—Theoretical models and perspectives. Functional Ecology, 33(1), 13–30. https://doi.org/10.1111/1365-2435.13241

Rees, M., & Ellner, S. P. (2016). Evolving integral projection models: Evolutionary demography meets eco-evolutionary dynamics. Methods in Ecology and Evolution, 7(2), 157–170. https://doi.org/10.1111/2041-210X.12487

Lande, R. (1982). A quantitative genetic theory of life history evolution. Ecology, 63(3), 607–615. http://www.primes.colostate.edu/Lande 1982.pdf

Rice, S. H. (2004). Evolutionary theory: mathematical and conceptual foundations. Sinauer Associates.

Gardner, A. (2020). Price’s equation made clear. Philosophical Transactions of the Royal Society B: Biological Sciences, 375(1797), 20190361. https://doi.org/10.1098/rstb.2019.0361

Gardner, A. (2008). The Price equation. Current Biology, 18(5), 198–202. https://doi.org/10.1016/j.cub.2008.01.005

Harper, J. L. (1967). A Darwinian Approach to Plant Ecology. In Source: Journal of Ecology (Vol. 55, Issue 2). https://about.jstor.org/terms

Pianka, E. R. (1974). Evolutionary ecology.

Roughgarden, J. (1979). Theory of Population Genetics and Evolutionary Ecology: An Introduction.

Begon, M. et al. (2020). Ecology: From Individuals to Ecosystems.

Hendry, A. P. (2017). Eco-evolutionary dynamics. Princeton university press.

Yamamichi, M., Ellner, S. P., & Hairston, N. G. (2023). Beyond simple adaptation: Incorporating other evolutionary processes and concepts into eco-evolutionary dynamics. Ecology Letters, 26(S1), S16–S21. https://doi.org/10.1111/ele.14197
